### 2020-06-17 paper/sex_selection_spacing-ver4.tex
Add sex selection definition and focus on questions

India has changed significantly over the last four decades:
the economy has grown substantially,
education levels have increased for both males and females,
and the total fertility rate has fallen to 2.2 
\citep{Bosworth2008,Dharmalingam2014,
International-Institute-for-Population-Sciences-IIPS2017}.

India has, however, also experienced a rapidly expanding 
access to prenatal sex determination.
Combined with continued strong son preference, the result has been a dramatic 
increase in the males-to-females ratio at birth
\citep{das_gupta97,Arnold2002,retherford03b,Guilmoto2012,
Portner2015b,Jayachandran2017}.

What the prior research on India has not appreciated is that since each abortion 
increases the interval between births by six to twelve months, the growing use of 
sex-selective abortions may substantially increase birth spacing.%
\footnote{
The increase consists of three parts.
First, starting from the abortion, the uterus needs at 
least two menstrual cycles to recover;  otherwise, the likelihood 
of spontaneous abortion increases substantially \citep{zhou00b}.
The second part is the waiting time to conception, which is between 
one and six months \citep{Wang2003}.
Finally, sex determination tests are reliable only from three months 
of gestation onwards.
}
Furthermore, the combination of increasing female education, 
lower fertility, higher household income, and the low and declining female 
labor force participation are also likely to impact birth spacing.
Hence, the changes in birth spacing may be much more substantial in India 
than what we have observed in other countries.

There are two primary motivations for examining birth spacing in this
situation, besides gaining a better general understanding of how
families make fertility decisions.

First, if birth intervals increased substantially over time, India's fertility may 
be higher than generally accepted based on the total fertility rate (TFR).
Increasing birth spacing is a tempo effect, which, by postponing births, makes the TFR a 
downward biased estimate of cohort fertility 
\citep{Hotz1997,Bongaarts1999,Ni-Bhrolchain2011}.

Second, longer birth spacing is associated with lower mortality and 
morbidity risk \citep{Conde-Agudelo2012,Molitoris2019}.
In India, the duration to the next birth has traditionally been shorter with fewer sons, 
which likely contributed to the higher mortality and morbidity risk for girls
\citep{Whitworth2002,Bhalotra2008,Maitra2008,Jayachandran2011,Jayachandran2017a}.
Longer spacing between births then potentially lowers mortality, although there may
be counteracting effects if the longer spacing arises because of multiple abortions.
There has, so far, been no attempt to examine how mortality has responded to the
increase in sex selection.

% Finally, birth spacing has long served as a measure of son preference, with strong
% son preference associated with shorter birth intervals, the fewer sons a family has 
% \citep{ben-porath76b,Leung1988}.
% Now, longer birth intervals in the absence of sons may instead arise because couples 
% with a strong preference for sons use sex selection, possibly to the point where we 
% observe a reversal of the traditional spacing pattern.
% Despite a large literature on son preference and birth spacing, no prior research has,
% however, examined how sex selection affects spacing.

I examine how birth spacing in India has changed over time and across 
groups with the introduction of sex selection.
Using data from the four National Family and Health Surveys (NFHS), I apply
a competing risk hazard model to the birth histories of Hindu women, covering
the period 1972 to 2016.
The competing risk hazard model directly incorporates the effects of 
sex-selective abortions on birth intervals
\emph{and} 
the likelihood of a son 
but also works in cases without prenatal sex selection.
The empirical model allows me to predict fertility, taking into
account both the likelihood of parity progression and the likelihood of having a child of 
a specific sex.
Finally, I examine how infant mortality risk has changed with the changing birth spacing 
and whether sex selection affects infant mortality.

>>>>

India saw many positive changes over the last four decades:
the economy grew substantially,
schooling increased for both males and females,
and the total fertility rate fell to 2.2 
\citep{Bosworth2008,Dharmalingam2014,
International-Institute-for-Population-Sciences-IIPS2017}.

However, the males-to-females ratio at birth also increased dramatically over this period
\citep{das_gupta97,Arnold2002,retherford03b,Guilmoto2012,Portner2015b,Jayachandran2017}.
Chorionic villus sampling (CVS), amniocentesis, and ultrasound all allow parents to learn 
a fetus' sex, and access to these technologies expanded rapidly.
With a continued strong preference for sons, the result has been an increase in sex 
selection, the selective abortion of female fetuses based on prenatal sex determination.%
\footnote{
CVS requires the shortest period of gestation---eight to twelve
weeks---which means that abortions can occur in the first trimester.
Amniocentesis can be performed after fourteen weeks but requires three
to four weeks before the result is available.
Hence, any abortion would not take place until more than halfway through the second 
trimester.
Ultrasound has the advantages of being non-invasive and relatively cheap.
The sex of a fetus can be determined in the third month of gestation if
it is a boy and the fourth month if it a girl.
}

The main question I examine is: how did birth spacing respond to these
significant changes, especially the introduction and spread of sex selection? 
This question is motivated by two factors. 
First, prior research has failed to appreciate that each abortion increases the
interval between births by six to twelve months.%
\footnote{
The increase consists of three parts. 
First, after an abortion, the uterus needs at least two menstrual cycles to recover; 
otherwise, the likelihood of spontaneous abortion increases substantially \citep{zhou00b}. 
The second part is the waiting time to conception, which is between one and six months. 
Finally, sex determination tests are reliable only from three months of gestation onwards.
}
Hence, the growing use of sex-selective abortions may substantially increase birth spacing, 
but we know little about how much. 
Second, more female education, higher household income, and low and declining female labor 
force participation are all likely to impact birth spacing, and the combined changes in 
birth spacing may, therefore, be much more substantial in India than what we have observed 
in other countries.

Understanding birth spacing is valuable not only because it contributes not only to our 
knowledge about fertility decisions but also because of the potential effects of birth 
spacing on outcomes such as fertility measures and mortality. 
I focus on two questions.

First, to what extent do changes in birth spacing bias our standard fertility estimates?
The tempo effect that arises from longer birth intervals means that the total fertility 
rate is a downward biased estimate of cohort fertility 
\citep{Hotz1997,Bongaarts1999,Ni-Bhrolchain2011}.
Hence, if birth intervals increased substantially over time, India's fertility may 
be higher and have declined slower than generally accepted.

Second, how do the changes in birth spacing and the increased use of abortions impact 
children's survival chances?
In India, the duration to the next birth has traditionally been shorter with fewer sons, 
which likely contributed to the higher mortality and morbidity risk for girls
\citep{Whitworth2002,Bhalotra2008,Maitra2008,Jayachandran2011,Jayachandran2017a}.
Longer birth spacing then potentially lowers mortality and morbidity risk 
\citep{Conde-Agudelo2012,Molitoris2019}.
There may, however, be counteracting effects if the longer spacing arises because of 
multiple abortions.
There has, so far, been no attempt to examine how infant mortality has responded 
to the increase in sex selection.

To examine how birth spacing in India has changed over time and across groups with the 
introduction of sex selection, I apply a competing risk hazard model to the birth 
histories of Hindu women, covering the period 1972 to 2016, using data from the four 
National Family and Health Surveys (NFHS).
The competing risk hazard model directly incorporates the effects of 
sex-selective abortions on birth intervals
\emph{and} 
the likelihood of a son 
but also works in cases without prenatal sex selection.
The empirical model allows me to predict fertility, taking into
account both the likelihood of parity progression and the likelihood of having a child of 
a specific sex.
I use the same data to examine how infant mortality risk has changed with the changing 
birth spacing and whether sex selection affects infant mortality.
